Likelihood-free Inference and Optimization \\ using Stochastic Gradient Approximations

### 2015-02-11 papers/uai-2015/draft.tex
ADAgrad RSMProp ADAm
sghmc and sg thermostats mostly

\newcommand{\gradestimate}[2]{\hat{g}_{#1}\lp#2\rp}
\newcommand{\gradestimaterepeat}[3]{\hat{g}^{#3}_{#1}\lp#2\rp}
\newcommand{\feval}[2]{y_{#1}^{#2}}
\newcommand{\cstep}[2]{c_{#1#2}}
\newcommand{\cstepconst}[1]{c_{#1}}
\newcommand{\dstep}[2]{\Delta_{#1#2}}
\newcommand{\dstepconst}[1]{\Delta_{#1}}
\newcommand{\loglik}[1]{L\lp#1\rp}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{STOCHASTIC-GRADIENT LD} \label{sgld}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

@@

>>>>

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{STOCHASTIC-GRADIENT LD} \label{sgld}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{eqnarray}
  h & = & \frac{N}{n} \sum_{i=1}^n \nabla \log p(\bx_i | \thetav)- \sum_{i=1}^N \nabla \log p(\bx_i | \thetav ) \\
    & = & \frac{N}{n} \sum_{i=1}^n s_i \\
  s_i & = & \nabla \log p(\bx_i | \thetav) + \frac{1}{N} \nabla p(\thetav) \\
  \text{V}(h) & = & \frac{N^2}{n^2} \sum_{i=1}^n \text{Var}(s_i) \\
                & = & \frac{N^2}{n^2} \sum_{i=1}^n \frac{1}{n} \lp \sum_{j=1} (s_i - \bar{s})^2 \rp \\
                & = & \frac{N^2}{n^2} \sum_{i=1}^n V_s \\
                & = & \frac{N^2}{n^2} n V_s \\
                & = & \frac{N^2}{n} V_s 
  %\text{Var}(\thetav) & = & \frac{\epsilon^2}{4}\text{Var}(h) + \epsilon \\
  %       \frac{\epsilon^2}{4}\text{Var}(h) & << & \epsilon \\
  %       \frac{\epsilon N^2}{4 n} V_s & << & 1       
\end{eqnarray}
\begin{eqnarray}
  \text{Var}(\thetav) & = & \frac{\epsilon^2}{4}\text{Var}(h) + \epsilon \\
         \frac{\epsilon^2}{4}\text{Var}(h) & << & \epsilon \\
         \frac{\epsilon N^2}{4 n} V_s & << & 1       
\end{eqnarray}

@@

\subsection{ABC Cost Function}

The ABC likelihood is proportional to the convolution between a kernel around the observation statistics $\y$ and the generator of pseudo-statistics $\x$, i.e. the simulator.  Since pseudo-statistics can only be generated, we resort to a Monte Carlo estimate for the ABC likelihood based on $S$ draws from the simulator using the same parameter vector $\thetav$ but different random seeds:
\begin{eqnarray}
  \pi\lp \y | \thetav \rp & = &  \int \pi_{\epsvec}\lp \y | \x \rp \pi\lp \x | \thetav \rp d \x \\
   & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | \x^{(s)} \rp
\end{eqnarray}
where $\x^{(s)} = \pi\lp \x | \thetav, \omega_s \rp$.  Note we have passed a unique random seed $\omega_s$ into the simulator, making $\x^{(s)}$ a deterministic function of the simulator with parameters $\{\thetav, \omega_s\}$.  It will be useful to rewrite this deterministic simulator function as $f$: $\x^{(s)} = f\lp \thetav, \omega_s \rp$, allowing us to rewrite the likelihood as
\begin{eqnarray}
  \pi\lp \y | \thetav \rp & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
\end{eqnarray}

For optimization purposes we can search for the maximum a posteriori (MAP) estimator $\hat{\thetav}$ using the log-joint distribution $\hat{\thetav} = \argmax_{\thetav} \mathcal{MAP}$, where $\mathcal{MAP}$ is
\begin{eqnarray}
  \mathcal{LJ} &    =    & \log \pi(\thetav) + \log \pi\lp \y | \thetav \rp \\
                & \approx &\log \pi(\thetav) + \log \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp - \log S
\end{eqnarray} 
we can also maximize the lower-bound of $\mathcal{LJ} \geq \mathcal{LB}$, where 
\begin{eqnarray}
  \mathcal{LB} & \approx &\log \pi(\thetav) + \sum_{s=1}^S \log  \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
\end{eqnarray} 
which will be more convenient for optimization for typical kernels (i.e. $\pi_{\epsilon}$ is Gaussian).

The simplest optimization is to set the objective function  $C$ to be $-\mathcal{LJ}$ or $-\mathcal{LB}$ and compute stochastic gradients using 2SPSA with $q$ repeats (negative log-likelihood because we will assume minimization):
\begin{eqnarray}
  \frac{\partial C}{\partial \thetav} & = & \frac{1}{q} \sum_{j=1}^{q} \frac{C\lp \thetav + c \Delta_j\rp - C{\lp\thetav - c \Delta_j\rp}}{2 c \Delta_j}
\end{eqnarray}
or 1SPSA
\begin{eqnarray}
  \frac{\partial C}{\partial \thetav} & = & \frac{1}{q} \sum_{j=1}^{q}  \frac{C\lp \thetav + c \Delta_j\rp - C(\thetav)}{c \Delta_j}
\end{eqnarray}
which is less accurate but is less computation if $C(\thetav)$ is computed at each step (TODO: possible algorithm).  TODO: rewrite $\frac{\partial C}{\partial \thetav}$ as $\hat{g}$ or similar.

\subsection{Special Case: Gaussian Kernels}
When the kernel is known we can use the chain rule to compute the gradients with the first part from the kernel and the second from the variation of the simulator pseudo-statistics, whose gradient can be estimated using SPSA.  Expanding the likelihood over $J$ statistics, the lower bound on the loglikelihood becomes:
\begin{eqnarray}
  \hat{L} & = & \sum_s \log \pi_{\epsilon}\lp \y | f(\thetav, \omega_s)\rp \\
          & = & \sum_s \log \prod_{j=1}^J \pi_{\epsilon_j}\lp y_j | f_j(\thetav, \omega_s)\rp \\
          & = & \sum_s \sum_j \log  \pi_{\epsilon_j}\lp y_j | f_j(\thetav, \omega_s)\rp \\
          & = & \sum_s \sum_j -\frac{1}{2} \frac{\lp y_j - f_j(\thetav, \omega_s)\rp^2}{\epsilon_j^2} - Z_j \\
\end{eqnarray}

The gradient:
\begin{eqnarray} 
  \nabla \hat{L} & = & \sum_s \sum_j \frac{\lp y_j - f_j(\thetav, \omega_s)\rp}{\epsilon_j^2} \cdot \frac{\partial f_j(\thetav, \omega_s)}{\partial \thetav} \\
                 & = & \sum_s \sum_j e_{js} \nabla_{js}
\end{eqnarray}
The first part $e_{js}$ is analytic, the second part $\nabla_{js}$ requires estimation by 2SPSA:
\begin{eqnarray} 
  \hat{ \nabla }_{js}  & = & \frac{1}{q} \sum_{r=1}^q \frac{f_j(\thetav + c \Delta_r , \omega_s) - f_j(\thetav - c \Delta_r , \omega_s)}{2 c \Delta_r}
\end{eqnarray}
or 1SPSA
\begin{eqnarray} 
  \hat{ \nabla }_{js} & = & \frac{1}{q} \sum_{r=1}^q \frac{f_j(\thetav + c \Delta_r , \omega_s) - f_j(\thetav, \omega_s)}{c \Delta_r}
\end{eqnarray}
the benefit of 1SPSA in this case is that the first gradients  $e_{js}$ can be computed from $f_j(\thetav, \omega_s)$.

\subsection{SGD, AdaGrad, RMSProp, AdaM}

With our stochastic gradient in hand, we can make use of robust stochastic gradient rules, including recent advances from machine learning.

\subsubsection{SGD}
The basic update rule is at time $t$ for minimization is
\begin{eqnarray}
  \Delta\thetav & = & - \alpha_t \hat{g_t}
\end{eqnarray} 
with conditions $\sum_{t=1}^{\infty} \alpha_t = \infty$ and $\sum_{t=1}^{\infty} \alpha_t^2 < \infty$ required for convergence.
An additional momentum parameter $\beta_1$ can be added to improve convergence
\begin{eqnarray}
  \gmom & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \Delta\thetav & = & - \alpha_t \gmom
\end{eqnarray}

\subsubsection{AdaGrad}
AdaGrad uses the total cumulative element-wise square of the gradients to automatically both decrease $\alpha_t$ and scale the learning rates for each dimension: 
\begin{eqnarray}
  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \gsquare & = & \sum_{t'=1}^t \hat{g_t}^2 \\
  \Delta\thetav & = & - \alpha_t \gmom / \sqrt{\gsquare}
\end{eqnarray}

\subsubsection{RMSProp}
\begin{eqnarray}
  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \gsquare & = & \beta_2 \gsquare + (1-\beta_2)*\hat{g_t}^2 \\
  \Delta\thetav & = & - \alpha_t \gmom / \sqrt{\gsquare}
\end{eqnarray}

\subsubsection{AdaM}
\begin{eqnarray}
  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
  \gsquare & = & \beta_2 \gsquare + (1-\beta_2)*\hat{g_t}^2 \\
  \gamma & = & \frac{ \sqrt{1-(1-\beta_2)^t}}{(1-(1-\beta_1)^t)} \\
  \Delta\thetav & = & - \alpha_t \gamma \gmom / \sqrt{\gsquare}
\end{eqnarray}

\section{Applying SGLD to ABC}

There are two main sources of randomness in the ABC gradient computation: the random perturbation mask and the random seed of the simulator model, which controls the simulator noise.  This noise can be eliminated by using common random numbers \cite{spall-crn}.  On the other hand, it is not clear that this is the best strategy, as the noise from the masks at a fixed seed may be higher than

\begin{eqnarray}
  V_{\gradestimaterepeat{t}{\thetav_t}{j}} & = & \frac{1}{q-1} \sum_{j=1}^q (\gradestimaterepeat{t}{\thetav_t}{j} - \gradestimate{t}{\thetav_t})^2 \\
  V_{\gradestimate{t}{\thetav_t}} & = & \frac{1}{q-1} V_{\gradestimaterepeat{t}{\thetav_t}{j}} \\
  \frac{\epsilon^2}{4 }V_{\gradestimate{t}{\thetav_t}} & = & \frac{\epsilon^2}{4 q }V_{\gradestimaterepeat{t}{\thetav_t}{j}} \\
                                                       & << & \epsilon \\
                    \frac{\epsilon}{4 (q-1) }V_{\gradestimaterepeat{t}{\thetav_t}{j}}  & << & 1                                  
\end{eqnarray}

Check for $g(\thetav) + h(\thetav) + \eta$ that $h(\thetav) + \eta \approx \eta$

Also $\theta_{t+1} = \theta_t + \frac{\epsilon}{2}\gradestimate{t}{\thetav_t} + \sqrt{\epsilon}\mathcal{N}(0,1)$

@@

From Byrd:
\begin{eqnarray}
\frac{||V_{\gradestimaterepeat{t}{\thetav_t}{j}} ||_1}{q} & \leq r^2 || \hat{g}_t ||_2^2 \\
\hat{q} & = & \frac{||V_{\gradestimaterepeat{t}{\thetav_t}{j}} ||_1}{r^2 || \hat{g}_t ||_2^2}
\end{eqnarray}

TODO:
\begin{itemize}
  \item Check that $q$ gradients are normal.  Especially as $q$ increases.
  \item Merge Byrd's test for $\hat{q}$ and condition on SGLD-ABC.
  \item Variance check for increasing $q$ can also be used to stop optimization since the variance is less than the noise in the gradients.
\end{itemize}

@@

\subsection{Optimization of Blowfly Dynamics}

\subsubsection{Using AdaM}

\begin{figure}[h]
\vspace{1in}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_series.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_series_sgld5k.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_sgld5k.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_series.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_series_sgld5k.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_sgld5k.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_noise.png}
\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_noise_sgld5k.png}
\caption{Optimizing Blowfly using $q=5$, $\epsilon = 0.01$, $\beta_1=0.75$,$\beta_2=0.9$,$\alpha=0.1$ and $c=1$.  Left: Optimization using AdaM, Right: ``SGLD'' with estimated noise. }
\end{figure}

\subsubsection{Using AdaM with ``SGLD" injected noise}
